---
author: "Prasaath Sastha"
date: "23 OCT 2024"
title: "Tidy Tuesday - 22OCT2024"
categories: [Tidy Tuesday, code, analysis]
code-fold: show
---

::: {.callout-note appearance="simple"}
This is my first tidy Tuesday challenge.
:::

# Introduction to Tidy Tuesday

This data is obtained from the [Tidy Tuesday repository hosted on github](https://github.com/rfordatascience/tidytuesday "link to the repository"). Side note: The tidy Tuesday repository gets updated with a new data set every Tuesday, and people all around the world use this data to produce some awesome visualization and shiny apps.

This is the first Dataset I am working on, and I am writing this quarto markdown as I go along. I have come across a lot of data scientists that do this in a R script to produce a plot or a graphic. And to me, I choose to do this in a markdown format, because I simply love the quarto platform. I am a big notion person, like half my life is on notion. So, using quarto with all the markdown and code execution functionality is like Disney land to me.

Apologies if I am boring you, since this is my first post I am just going over what happening in my head at this time. This blabber will not be present on the future posts.

So lets get started, the data we are going to read and work on are released on [22 OCT 2024](https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-10-22).

The libraries that I would like to use on any R project for efficiency are

```{r}
#| code-fold: show
#| warning: false
#| label: libraries
# Load libraries

library(tidyverse) # This is a collection of packages that is used for data exploration, wrangling and plotting.
library(tidytuesdayR) # This package is used to connect to the tidy Tuesday repo for the data
library(knitr)
# Library that plays a hige role in improving test plot aesthetics
library(patchwork)
library(ggrepel)
library(glue) 
library(ggtext) 
library(showtext)

# Libraries for Maps
library(tmap)
library(sf)
library(rnaturalearth)
```

Now the `tidytuesdayR` package helps us access the data set that is present in the git hub repo.

```{r}
#| label: tbl-glimpse
#| warning: false
tuesdata <- tt_load("2024-10-22")

cia_factbook <- tuesdata$cia_factbook

glimpse(cia_factbook)

```

Using glimpse we can see that the data set has 259 observations over 11 variables. On a superficial view, it appears that the data set contains variables such as area, birth rate, death rate, and so on with respect to each country.

Now, since we have come to an understanding that each of the variable is associated with a country, we can go on start making basic structure for how we are going to analyze or visualize the data.

After this step, I would like to examine each of the variable by summarizing or plotting to see if there are any null values or extreme outliers that needs to be evaluated further. Since all the variables except the country are numbers or doubles there are a list of things I like to check before proceeding

-   Missing values and their representation
-   Negative values (appropriateness)
-   Minimum and Maximum (context specific)
-   Data format

If it is a categorical variable, i always check for duplicates, and similar names.

#### Data Dictionary from the CIA factbook data set

| variable | class | description |
|------------------------------|-----------------------|-----------------------------------------------------------|
| country | integer | Name of the country (factor with 259 levels). |
| area | integer | Total area of the country (in square kilometers). |
| birth_rate | double | Birth rate (number of live births per 1,000 people). |
| death_rate | double | Death rate (number of deaths per 1,000 people). |
| infant_mortality_rate | double | Infant mortality rate (number of deaths of infants under one year old per 1,000 live births). |
| internet_users | integer | Number of internet users. |
| life_exp_at_birth | double | Life expectancy at birth (in years). |
| maternal_mortality_rate | integer | Maternal mortality rate (number of maternal deaths per 100,000 live births). |
| net_migration_rate | double | Net migration rate (number of migrants per 1,000 people). |
| population | integer | Total population of the country. |
| population_growth_rate | double | Population growth rate (multiplier). |

```{r}
#| label: tbl-summary
summarytable <- summary(cia_factbook)

kable(summarytable)

```

```{r}
# Creating some general variables for names and titles for axis and legenda

title_area <- "Area"
title_birth_rate <- "Birth Rate"
title_death_rate <- "Death Rate"
title_infantMR <- "Infant Mortality Rate"
title_maternalMR <- "Maternal Mortality Rate"
title_interetuser <- "Internet Users"
title_netmigration <- "Net Migration Rate"
title_population <- "Population"

t_population_growth_rate <- tibble(title = "Population Growth Rate")
t_area <- tibble(title = "Area", unit = "Sq Km")
t_birth_rate <- tibble(title = "Birth Rate", unit = "live births per 1,000 people")
t_death_rate <- tibble(title = "Death Rate", unit = "deaths per 1,000 people")
t_infantMR <- tibble(title = "Infant Mortality Rate", unit = "deaths of infants under one year old per 1,000 live births")
t_maternalMR <- tibble(title = "Maternal Mortality Rate", unit = "maternal deaths per 100,000 live births")
t_interetuser <- tibble(title = "Internet Users", unit = "n")
t_netmigration <- tibble(title = "Net Migration Rate", unit = "migrants per 1,000 people")
t_population <- tibble(title = "Population", unit = "n")
```

```{r}
#| label: fig-boxplots
#| fig-height: 10
#| fig-width: 15
boxplot <- function(data, var, xlab = "X-Axis", ylab = "Y-Axis", main = "My Scatter Plot", col = "blue") {
  # Create a dataframe from x and y

  # Generate the plot
  ggplot(data = data, aes(x = {{var}})) +
    geom_boxplot(color = col, outlier.colour = "red", na.rm = TRUE, orientation = "y")+
    scale_y_continuous(labels = NULL)+
    labs(x = xlab, y = ylab)
}


A <- boxplot(cia_factbook, area, xlab = paste(t_area$title, "(", t_area$unit, ")"), ylab = "")
B <- boxplot(cia_factbook, birth_rate , xlab = paste(t_birth_rate$title, "(", t_birth_rate$unit, ")"), ylab = "")
C <- boxplot(cia_factbook, death_rate, xlab = paste(t_death_rate$title, "(", t_death_rate$unit, ")"), ylab = "")
D <- boxplot(cia_factbook, infant_mortality_rate , xlab = paste(t_infantMR$title, "(", t_infantMR$unit, ")"), ylab = "")
E <- boxplot(cia_factbook, internet_users , xlab = paste(t_interetuser$title, "(", t_interetuser$unit, ")"), ylab = "")
F <- boxplot(cia_factbook, maternal_mortality_rate , xlab = paste(t_maternalMR$title, "(", t_maternalMR$unit, ")"), ylab = "")
G <- boxplot(cia_factbook, net_migration_rate , xlab = paste(t_netmigration$title, "(", t_netmigration$unit, ")"), ylab = "")
H <- boxplot(cia_factbook, population , xlab = paste(t_population$title, "(", t_population$unit, ")"), ylab = "")
I <- boxplot(cia_factbook, population_growth_rate , xlab = t_population_growth_rate$title, ylab = "")


A+B+C+D+E+F+G+H+I+ plot_layout(ncol = 2)
```

From the above series of plots we can see there are a number of outliers in each variable, they can be investigated individually by checking what country those variables are at the extreme ends and check them for coherence with real world knowledge.

And there are also missing values in all of the variables, which can be seen from the summary tables. Since its listed as NA's we are going to assume that the relevant data is not available for those variables. Generally, I would investigate the root-cause for missing data and the outliers, but for this tidy Tuesday visualization, I am willing to assume that these are genuine representations of the data.

Since the data are country specific, the following visualizations are plotted on maps. Some of the popular libraries for plotting data on maps are `sf`, `rnaturalearth`, and `tmap` . The following code creates a function that can be reused to plot different variables without having to redefine the plotting parameters each time. \

```{r}
#| label: fig-mapplots
world <- ne_countries(scale = "medium", returnclass = "sf")

cia_world_data <- world |> 
  left_join(cia_factbook, by = c("name" = "country"))

map_cia <- function(data, var, title, value) {
  ggplot(data = data) +
    geom_sf(aes(fill = {{ var }}), color = "black", size = 0.1) + 
    scale_fill_continuous(type = "viridis", na.value = "grey90", name = value, limits = NULL) +
    labs(title = title, caption = "Countries with no data are shaded in grey") +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      legend.text.position = "bottom",
      legend.title.position = "top",
      legend.key.width = unit(2, "cm")
    )
}

map_cia(cia_world_data, internet_users, title= t_interetuser$title, value = t_interetuser$unit)
map_cia(cia_world_data, population, title = t_population$title, value = t_population$unit)
map_cia(cia_world_data, birth_rate, title = t_birth_rate$title, value = t_birth_rate$unit)
map_cia(cia_world_data, death_rate, title = t_death_rate$title, value = t_death_rate$unit)
map_cia(cia_world_data, infant_mortality_rate, title = t_infantMR$title, value = t_infantMR$unit)
map_cia(cia_world_data, maternal_mortality_rate, title = t_maternalMR$title, value = t_maternalMR$unit)
map_cia(cia_world_data, population_growth_rate, title = t_population_growth_rate$title, value = "")
map_cia(cia_world_data, net_migration_rate, title = t_netmigration$title, value = t_netmigration$unit)

```

After seeing some general information of each variables on the above plots, now lets focus on the main question of the tidy Tuesday challenge

::: callout-note
Which countries have the highest number of internet users per square kilometer?

Which countries have the highest percentage of internet users?
:::

To answer the above questions, new variables need to be computed. For the first question, the computed variable is "internet_users_per_square_Km", and for the second one, "percentage_internet_users"

```{r}
#| label: fig-ttanswer
#| 

top10 <- cia_factbook |> 
  mutate(
    internet_users_per_square_Km = internet_users/area,
    percent_internet_users = (internet_users/population)*100,
    .keep = "all"
  )

p1 <- top10 |> arrange(desc(internet_users_per_square_Km)) |> slice_head(n=10) |> ggplot(aes(x = reorder(country, internet_users_per_square_Km), y = internet_users_per_square_Km, fill = country))+
  geom_bar(stat = "identity", show.legend = FALSE)+
  coord_flip()+
  labs(
    title = "Top 10 countries with most internet users per square Kilometer",
    y = "Number of internet users per square Kilometers",
    x = NULL
  )+
  geom_text(
    aes(label = round(internet_users_per_square_Km, 2)),
    position = position_dodge(width = 0.5)
  )+
  theme_minimal()
  
p2 <- top10 |> arrange(desc(percent_internet_users)) |> slice_head(n=10) |> ggplot(aes(x = reorder(country, percent_internet_users), y = percent_internet_users, fill = country))+
  geom_bar(stat = "identity", show.legend = FALSE)+
  coord_flip()+
  labs(
    title = "Top 10 countries with most internet users",
    y = "Percentage of Internet Users",
    x = NULL
  )+
  geom_text(
    aes(label = round(percent_internet_users, 2)),
    position = position_dodge(width = 0.5)
  )+
  theme_minimal()

p1
p2

cia_internet_data <- world |> 
  left_join(top10, by = c("name" = "country"))

map_cia(cia_internet_data, internet_users_per_square_Km, title = "Internet users per square Kilometer", value = "Number of People")
map_cia(cia_internet_data, percent_internet_users, title = "Percentage of Internet Users", value = "%")

```

To answer the questions for this tidy tuesday challenge we have computed 2 variables using the data provided in the CIA fact book. From which we can see the top 10 countries with highest number of internet users per square kilometer and countries with highest internet user percentage.
